Outliers Detection and Comparison of Origin-Destination Flows with Data Depth

### 2018-10-09 giscience_v2/giscience/giscience2018_old.tex
xx

%% Please provide for each author the \author and \affil macro, even when authors have the same affiliation, i.e. for each author there needs to be the  \author and \affil macros
\author[1]{Myeong-Hun Jeong}
\author[2]{Junjun Yin}
\author[3]{Shaowen Wang}
\affil[1]{Department of Civil Engineering, Chosun University, Gwangju, Republic of Korea\\
  \texttt{mhjeong@chosun.ac.kr}}
\affil[2]{Social Science Research Institute; Institute for CyberScience, Penn State University, PA, USA\\
  \texttt{jyin@psu.edu}}
\affil[3]{Departmet of Geography and Geographic Information Science, University of Illinois at Urbana-Champaign, IL, USA\\
	\texttt{shaowen@illinois.edu}}
\authorrunning{M.-H. Jeong, J. Yin and S. Wang} %mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et. al.'

@@

\begin{abstract}
The advances in location-aware technologies have generated a huge volume of trajectory data. In particular, Origin-Destination (OD) trajectories provide rich information to understand urban flow and transport demand. This study presents a new methodology to detect OD flows outliers and conduct hypothesis testing between two OD flows in terms of the variations of spatial extent (i.e., spread). The proposed method is based on data depth, which measures the centrality and outlyingness of a point with respect to a given data set in $\mathbb{R}^d$. Based on the center-outward ordering property, it is possible to analyze the underlying OD flows characteristics such as location, outlyingness, and  spread. The proposed method is compared with Mahalanobis distance approach to detect OD anomalies, and F-test to identify the difference in scale. Empirical evaluation has demonstrated that the proposed method can identify OD flows outliers in an interactive way. Further, it can provide new perspectives by considering the overall structure of data when comparing two different OD flows in scale.

@@

\section{Introduction}
With the rapid rise in ubiquity of geolocation-aware sensors, knowledge discovery is greatly enhanced by extracting and mining interesting patterns from spatiotemporal big data in various domains.
In particular, the location-acquisition technologies generate large volumes of movement data, which are used to track people, animals, vehicles, and even natural phenomena.
Such data help us better model moving objects and reveal hidden patterns that are important to urban planning and its applications in characterizing urban human mobility \cite{yin2017depicting}, accessibility of the activity space \cite{kwan1998space}, and sustainability of the urban systems \cite{alberti2003integrating}.
Importantly, trajectory mining leads to solutions for addressing important research problems in different fields, such as urban planning \cite{mazimpaka15AGILE}, transportation \cite{chen13Percom}, environment \cite{devarakonda13SIGKDD}, and public security and safety \cite{buchin14JOSIS}.

@@

In recent years, researchers have investigated a variety of approaches to trajectory data mining.
In general, the contemporary trajectory mining methods can be classified into four categories: clustering, classification, frequent/group pattern mining, and outlier detection \cite{mazimpaka16JOSIS,zheng15ACMTIST}. These techniques can be used independently or combinatorially for trajectory mining applications. For this study, we focus on outlier detection of OD flows. Outlier detection aims to find trajectories that do not follow the typical flow of trajectory data sets for characterizing the connectivity between regions \cite{mazimpaka16JOSIS}. \cite{fontes13GeoInfo,lee08ICDE} use Euclidean distance to find outlier patterns from trajectories. \cite{pan13ACMGIS,liu12IJGIS} raise the questions of Euclidean distance approach due to the loss of local features and unavailability when external factors affect the trajectories (e.g., topography, land cover or weather condition). \cite{pan13ACMGIS,liu12IJGIS} addresses this issue by using robust distance measurements (i.e., Mahalanobis distance \cite{pan13ACMGIS} and relative distance \cite{liu12IJGIS}). Further, structural features \cite{yuan11JCIS} and a data-induced random tree \cite{zhang11UC} are exploited to detect anomalous trajectories instead of using distance or density. In a similar fashion, this study considers the center regions of OD flows with data depth in order to robustly detect the OD flow outliers.

In addition, utilizing visual analytics is a common approach to analyze OD flow data.
Visual representation of massive movement data enables comprehensive exploration of data and results in understanding complex flow trends.
Aggregation and generalization of movement data are frequently utilized \cite{andrienko08VAST,adrienko11IEEETVCG,guo14IEEETVCG}.
While visual analytics can help extract inherent patterns from massive data, it is difficult to quantitatively compare two sets of different OD flows based on a hypothesis testing.
In other words, it is complicated to comprehend how two OD flows differ and, more importantly, by how much.
In this light, this paper uses bivariate hypothesis testing methods based on data depth to understand how different two OD flow data sets are in terms of the amount of spatial extent.

It is worth noting that flow mapping approaches frequently suffer from the modifiable areal unit problem (MAUP). For instance, it is not guaranteed that different aggregations via location can present coherent patterns.
Kernel-based flow estimation and smoothing are used to overcome different spatial resolution \cite{guo14IEEETVCG}. 
Instead of focusing on finding the best areal unit to partitioning the urban space and aggregating the OD flows, this study chooses traffic analysis zones in New York City as a base unit. However, the method can be adapted to other choices of areal units.
In this study, New York City taxi trip data include the origin and destination within traffic analysis zones, which ignores the intermediate locations of the actual routes.
It is not necessary to reconstruct individual movements for flow estimation (see \cite{duckham16ICGIS}).

@@

The bag plot is able to present the location (the depth median), spread (the spatial extent of bag), correlation (the orientation of the bag), and skewness (the shape of the bag and the fence) of data \cite{rousseeuw99AS}. For example, we can see which forward OD flows have higher counts compared to the reverse OD flows; the relatively linear correlation between forward OD flows and reverse OD flows; and the skewness of forward (reverse) OD flows in Figure \ref{fig:OD_0521}.

@@

In addition, it is apparent that the bootstrap method is a time consuming process. We generate $2,000$ bootstrap samples. In order to scale up the bootstrap computation, we distribute work across multiple nodes and cores by implementing an embarrassingly parallel R code.  Thus, we can effectively improve execution efficiency.
% investigate resource pressures between serial and parallel approaches in the following section.

@@

As a case study, this study used OD flows during weekdays and weekends in June, 2014. The weekdays data set includes taxi trajectories on June 3, 10, 17, and 24. There are 1,721,655 taxi trips. The weekends data set includes taxi trajectories on June 8, 15, 22, and 29.  Its taxi trips are 1,593,480.

\subsection{Procedure}
The performance of the proposed method was compared with alternative methods. Trajectories anomalies detection, based on Mahalanobis distance \cite{pan13ACMGIS},  was used to compare the performance of outliers detection. The Mahalanobis distance considers the correlations of the data (e.g., two OD flows), which distinguishes it from Euclidean distance. According to \cite{pan13ACMGIS}, the anomaly detection threshold can be defined as below.

@@

With regard to data cleaning process, this study used Hadoop with Pig. We have developed a Hadoop program to handle the large volume of data (~173 million taxi trip records), remove trips with invalid OD coordinates, and assign each OD locations into the corresponding traffic analysis zone.  
R was used to implement the OD flow outlier detection. The computing environment uses Amazon Web Service and Bridges supercomputer at the Pittsburgh Supercomputing Center. Further, we used OD flows that have more than 10 trips. It is reasonable to remove the majority of low number of OD flows, which may distort the view of the data. All the code will be released as open source (the link to the code will added later and currently is available upon request).

@@

\begin{figure}
	\centering
	\begin{subfigure}[b]{0.49\textwidth}
		\includegraphics[width=\textwidth]{images/Outliers_rare_weekdays_weekends.eps}
		\caption{Relative OD flows outliers}
		\label{fig:weekdays_rare}
	\end{subfigure}
	\hfill %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 	%(or a blank line to force the subfigure onto a new line)
	\begin{subfigure}[b]{0.49\textwidth}
		\includegraphics[width=\textwidth]{images/outliers_rare2_weekdays_weekends.eps}
		\caption{OD flow map for relative OD flows outliers}
		\label{fig:weekdays_rare_map}
	\end{subfigure}
	\caption{Relative OD flows outliers on weekdays and weekends: Triangles in Figure \ref{fig:weekdays_rare} coincide with red lines in Figure \ref{fig:weekdays_rare_map}. }\label{fig:weekdays_rare_OD_map}	
\end{figure}

@@

\begin{figure}
	\centering
	\begin{subfigure}[b]{0.49\textwidth}
		\includegraphics[width=\textwidth]{images/out_weekdays_weekends_md2_outlier.eps}
	\end{subfigure}
	\caption{OD flows outliers on weekdays and weekends based on Mahalanobis distance. }\label{fig:week_ends_MD}	
\end{figure}

>>>>

%% Please provide for each author the \author and \affil macro, even when authors have the same affiliation, i.e. for each author there needs to be the  \author and \affil macros
\author[1]{Myeong-Hun Jeong}
\author[2]{Junjun Yin}
\author[3]{Shaowen Wang}
\affil[1]{Department of Civil Engineering,  Gwangju, Republic of Korea\\
  \texttt{mhjeong@chosun.ac.kr}}
\affil[2]{Social Science Research Institute, Penn State University, PA, USA\\
  \texttt{jyin@psu.edu}}
\affil[3]{Department of Geography and Geographic Information Science, the University of Illinois at Urbana-Champaign, IL, USA\\
	\texttt{shaowen@illinois.edu}}
\authorrunning{M.-H. Jeong, J. Yin and S. Wang} %mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et. al.'

@@

\begin{abstract}
The advances in location-aware technologies have generated a huge volume of trajectory data. In particular, Origin-Destination (OD) trajectories provide rich information to understand urban flow and transport demand. This study presents a new methodology to detect OD flows outliers and conduct hypothesis testing between two OD flows in terms of spread. The proposed method is based on data depth, which measures the centrality and outlyingness of a point with respect to a given data set in $\mathbb{R}^d$. Based on the center-outward ordering property, it is possible to analyze the underlying OD flows characteristics such as location, outlyingness, and  spread. The proposed method is compared with Mahalanobis distance approach to detect OD anomalies, and F-test to identify the difference in scale. Empirical evaluation has demonstrated that the proposed method can identify OD flows outliers in an interactive way. Further, it can provide new perspectives by considering the overall structure of data when comparing two different OD flows in scale.

@@

\section{Introduction}
With the rapid rise in ubiquity of geolocation-aware sensors, knowledge discovery is greatly enhanced by extracting and mining interesting patterns from spatiotemporal big data in various domains.
In particular, the location-acquisition technologies generate large volumes of movement data, which are used to track people, animals, vehicles, and even natural phenomena.
Such data help us better model moving objects and reveal hidden patterns that are important to urban planning and its applications in characterizing urban human mobility \cite{jiang2012discovering}, accessibility of the activity space \cite{kwan1998space}, and sustainability of the urban systems \cite{alberti2003integrating}. 
Importantly, trajectory mining leads to solutions for addressing important research problems in different fields, such as urban planning \cite{mazimpaka15AGILE}, transportation \cite{chen13Percom}, environment \cite{devarakonda13SIGKDD}, and public security and safety \cite{buchin14JOSIS}.

@@

In recent years, researchers have investigated a variety of approaches to trajectory data mining.
In general, the contemporary trajectory mining methods can be classified into four categories: clustering, classification, frequent/group pattern mining, and outlier detection \cite{mazimpaka16JOSIS,zheng15ACMTIST}. These techniques can be used independently or combinatorially for trajectory mining applications. For this study, we focus on outlier detection of OD flows. Outlier detection aims to find trajectories that do not follow the typical flow of trajectory data sets for characterizing the connectivity between regions \cite{mazimpaka16JOSIS}. \cite{fontes13GeoInfo,lee08ICDE} use Euclidean distance to find outlier patterns from trajectories. \cite{pan13ACMGIS,liu12IJGIS} raise the questions of Euclidean distance approach due to the loss of local features and unavailability when external factors affect the trajectories (e.g., topography, land cover or weather condition).  \cite{pan13ACMGIS,liu12IJGIS} addresses this by using robust distance measurements (i.e., Mahalanobis distance \cite{pan13ACMGIS} and relative distance \cite{liu12IJGIS}). Further, structural features \cite{yuan11JCIS} and a data-induced random tree  \cite{zhang11UC} are exploited to detect anomalous trajectories instead of using distance or density. In a similar fashion, this study considers the center regions of OD flows with data depth in order to robustly detect the trajectory outliers.

In addition, utilizing visual analytics is a common approach to analyze OD flow data.
Visual representation of massive movement data enables comprehensive exploration of data and results in understanding complex flow trends.
Aggregation and generalization of movement data are frequently utilized \cite{andrienko08VAST,adrienko11IEEETVCG,guo14IEEETVCG}.
While visual analytics can help to extract inherent patterns from massive data, it is difficult to quantitatively compare two sets of different OD flows based on a hypothesis testing.
In other words, it is complicated to comprehend how two OD flows differ and, more importantly, by how much.
In this light, this paper uses bivariate hypothesis testing methods based on data depth to understand how different two OD flow data sets are in terms of the amount of spatial extent.

It is worth noting that flow mapping approaches frequently suffer from the modifiable areal unit problem (MAUP). For instance, it is not guaranteed that different aggregations via location can present coherent patterns.
Kernel-based flow estimation and smoothing are used to overcome different spatial resolution \cite{guo14IEEETVCG}. 
Instead of focusing on finding the best areal unit to partitioning the urban space and aggregating the OD flows, this study chooses traffic analysis zones in New York City as a base unit. However, the method can be adapted to other choices of areal units.
In addition, New York City taxi trip data include the origin and destination within traffic analysis zones, which ignores the intermediate locations of the actual routes.
It is not necessary to reconstruct individual movements for flow estimation (see \cite{duckham16ICGIS}).

@@

The bag plot is able to present the location (the depth median), spread (the size of bag), correlation (the orientation of the bag), and skewness (the shape of the bag and the fence) of data \cite{rousseeuw99AS}. For example, we can see which forward OD flows have higher counts compared to the reverse OD flows; the relatively linear correlation between forward OD flows and reverse OD flows; and the skewness of forward (reverse) OD flows in Figure \ref{fig:OD_0521}.

@@

In addition, it is apparent that the bootstrap method is a time consuming process. We generate $2,000$ bootstrap samples. In order to scale up the bootstrap computation, we spread work across multiple nodes and cores by implementing an embarrassingly parallel R code.  Thus, we can effectively improve  execution times.
% investigate resource pressures between serial and parallel approaches in the following section.

@@

As a case study, this study used OD flows during weekdays and weekends in June. The weekdays data set includes taxi trajectories on June 3, 10, 17, and 24. There are 1,721,655 taxi trips. The weekends data set includes taxi trajectories on June 8, 15, 22, and 29.  Its taxi trips are 1,593,480.

\subsection{Procedure}

The performance of the proposed method was compared with alternative methods. Trajectories anomalies detection, based on Mahalanobis distance \cite{pan13ACMGIS},  was used to compare the performance of outliers detection. The Mahalanobis distance considers the correlations of the data (e.g., two OD flows), which distinguishes it from Euclidean distance. According to \cite{pan13ACMGIS}, the anomaly detection threshold can be defined as below.

@@

With regard to data cleaning process, this study used Hadoop with Pig. R was used to implement the code in Amazon Web Service  and Bridges supercomputer resources. Further, we used OD flows that have more than 10 trips. It is reasonable to remove the majority of low number of OD flows, which may distort the view of the data.

@@

\begin{figure}
	\centering
	\begin{subfigure}[b]{0.49\textwidth}
		\includegraphics[width=\textwidth]{images/Outliers_rare_weekdays_weekends.eps}
		\caption{Relative OD flows outliers}
		\label{fig:weekdays_rare}
	\end{subfigure}
	\hfill %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 	%(or a blank line to force the subfigure onto a new line)
	\begin{subfigure}[b]{0.49\textwidth}
		\includegraphics[width=\textwidth]{images/outliers_rare_weekdays_weekends.png}
		\caption{OD flow map for relative OD flows outliers}
		\label{fig:weekdays_rare_map}
	\end{subfigure}
	\caption{Relative OD flows outliers on weekdays and weekends: Triangles in Figure \ref{fig:weekdays_rare} coincide with red lines in Figure \ref{fig:weekdays_rare_map}. }\label{fig:weekdays_rare_OD_map}	
\end{figure}

@@

\begin{figure}
	\centering
	\begin{subfigure}[b]{0.49\textwidth}
		\includegraphics[width=\textwidth]{images/out_weekdays_weekends_md2_outlier.png}
	\end{subfigure}
	\caption{OD flows outliers on weekdays and weekends based on Mahalanobis distance. }\label{fig:week_ends_MD}	
\end{figure}
